---
title: "01_data_preprocessing"
author: "Hugo Maruyama"
date: "2026-03-06"
format:
  html:
    toc: true
    embed-resources: true
execute:
  warning: false
  message: false
---

## Introduction

This document details the preprocessing workflow for 16S rRNA gene sequencing data generated from both Illumina (MiSeq) and Oxford Nanopore Technologies (ONT) platforms. The primary objective is to import, clean, and harmonize taxonomic data from disparate pipelines to ensure robust comparability in downstream analyses.

## Environment Setup

Core R packages required for data manipulation, efficient file import, and academic-grade reporting are loaded.

```{r}
#| label: setup
library(tidyverse)
library(vroom)
library(knitr)
```

## Configuration

Input paths for the abundance tables are defined below. These datasets represent the outputs of platform-specific bioinformatics workflows: a QIIME2-based pipeline for Illumina data and the EPI2ME `wf-16s` workflow for Nanopore data.

**Note:** The file paths provided below are specific to the local directory structure of this project. Users should modify these paths to match their own environment and data locations.

```{r}
#| label: configuration

# Path to the Illumina (MiSeq) species-level abundance table (TSV)
illumina_input_path <- "./miseq/table_species_V3-V4_relative/table_species_V3-V4_relative.tsv"

# Path to the Nanopore (ONT) species-level abundance table (TSV)
nanopore_input_path <- "./epi2melabs/instances/wf-16s_01K5Z7KHWZP8KBZPWVWT3QSX91/output/abundance_table_species.tsv"
```

## Data Processing

### Illumina (MiSeq) Data Processing

The Illumina data is imported, and columns are renamed to match experimental sample identifiers. Relative abundances are expressed as percentages rounded to three decimal places. The taxonomic string is then parsed into standard hierarchical levels, and platform-specific prefixes (e.g., `k__`, `p__`) are removed to normalize the nomenclature.

```{r}
#| label: process-illumina

# Mapping internal identifiers to sample labels
illumina_samples <- c(AP = "1749", BP = "1750", AS = "1751", BS = "1752")
tax_levels <- c("kingdom", "phylum", "class", "order", "family", "genus", "species")

# Import, rename, and normalize Illumina data
illumina_processed <- vroom::vroom(
  file = illumina_input_path,
  delim = "\t", 
  skip = 1,
  show_col_types = FALSE
) %>%
  dplyr::rename(taxonomy = 1) %>%
  dplyr::select(taxonomy, !!!illumina_samples) %>%
  # Convert fractional abundance to rounded percentage (%)
  dplyr::mutate(dplyr::across(
    .cols = where(is.numeric),
    .fns = ~ round(.x * 100, 3)
  )) %>%
  # Parse semicolon-delimited taxonomy string
  tidyr::separate(
    col = taxonomy,
    into = tax_levels,
    sep = ";",
    fill = "right"
  ) %>%
  # Remove taxonomic prefixes and convert empty strings to NA
  dplyr::mutate(
    dplyr::across(
      .cols = dplyr::all_of(tax_levels), 
      .fns = ~ stringr::str_remove(., "^[kpcofgsd]?__")
    ),
    dplyr::across(
      .cols = dplyr::all_of(tax_levels), 
      .fns = ~ dplyr::na_if(., "")
    )
  )

# Preview the processed Illumina dataset
kable(head(illumina_processed), caption = "Processed Illumina Taxonomic Table")
```

### Nanopore (ONT) Data Processing

The Nanopore pipeline involves converting raw read counts to relative percentages and refining taxonomic labels. This includes removing intermediate "sub-kingdom" levels and standardizing placeholder taxon names (e.g., "oral taxon") to the Human Oral Microbiome Database (HOMD) "HMT" nomenclature to ensure cross-platform consistency.

```{r}
#| label: process-nanopore

# Mapping ONT barcodes to experimental groups
nanopore_samples <- c(AP = "barcode07", BP = "barcode08", AS = "barcode09", BS = "barcode10")
temp_tax_levels <- c("kingdom", "sub_kingdom", "phylum", "class", "order", "family", "genus", "species")
final_tax_levels <- c("kingdom", "phylum", "class", "order", "family", "genus", "species")

# Import and normalize Nanopore data
nanopore_processed <- vroom::vroom(
  file = nanopore_input_path,
  delim = "\t",
  show_col_types = FALSE) %>%
  dplyr::rename(taxonomy = 1) %>%
  dplyr::select(taxonomy, !!!nanopore_samples) %>%
  # Calculate relative abundance (%) from raw read counts
  dplyr::mutate(dplyr::across(
    .cols = where(is.numeric),
    .fns = ~ round(.x / sum(.x) * 100, 3)
  )) %>%
  # Separate taxonomy and retain standard hierarchical levels
  tidyr::separate(col = taxonomy, into = temp_tax_levels, sep = ";", fill = "right") %>%
  dplyr::select(dplyr::all_of(final_tax_levels), dplyr::all_of(names(nanopore_samples))) %>%
  # Standardize "Unclassified" and "Unknown" entries to NA
  dplyr::mutate(dplyr::across(
    .cols = dplyr::all_of(final_tax_levels),
    .fns = ~ dplyr::if_else(tolower(.x) %in% c("unclassified", "unknown"), NA_character_, .x)
  )) %>%
  # Refine species naming by removing redundant genus prefixes
  dplyr::mutate(
    species = dplyr::if_else(
      !is.na(genus) & stringr::str_starts(species, paste0(genus, " ")),
      stringr::str_remove(species, paste0("^", genus, " ")),
      species
    )
  ) %>%
  # Address species-level "Unclassified" labels
  dplyr::mutate(
    species = dplyr::if_else(stringr::str_starts(species, "Unclassified"), NA_character_, species)
  ) %>% 
  # Map "oral taxon" descriptors to standardized HMT identifiers
  dplyr::mutate(
    species = stringr::str_replace(species, "oral taxon (\\d+)", "HMT-\\1")
  )

# Preview the processed Nanopore dataset
kable(head(nanopore_processed), caption = "Processed Nanopore Taxonomic Table")
```

## Taxonomic Harmonization

To ensure comparability between the two platforms, manual and lineage-based harmonization is performed to correct discrepancies arising from the use of different reference databases (e.g., NCBI vs. HOMD).

### Targeted Taxonomic Corrections

Specific rules are defined to align the nomenclature, such as the classification of *Campylobacter* within the *Campylobacterota* phylum, reflecting current taxonomic standards.

```{r}
#| label: tax-harmonization-manual

# Define specific correction rules for harmonization
tax_correction_map <- tibble::tribble(
  ~condition_column, ~condition_value, ~target_column, ~correct_value,
  "genus", "Campylobacter", "phylum", "Campylobacterota"
)

# Apply corrections iteratively to both datasets
for(i in 1:nrow(tax_correction_map)) {
  rule <- tax_correction_map[i, ]
  
  illumina_processed <- illumina_processed %>%
    dplyr::mutate(
      !!rule$target_column := dplyr::if_else(
        !is.na(!!rlang::sym(rule$condition_column)) & !!rlang::sym(rule$condition_column) == rule$condition_value,
        rule$correct_value,
        !!rlang::sym(rule$target_column)
      )
    )
  
  nanopore_processed <- nanopore_processed %>%
    dplyr::mutate(
      !!rule$target_column := dplyr::if_else(
        !is.na(!!rlang::sym(rule$condition_column)) & !!rlang::sym(rule$condition_column) == rule$condition_value,
        rule$correct_value,
        !!rlang::sym(rule$target_column)
      )
    )
}
```

### Advanced Lineage Correction (Patescibacteria/TM7)

Taxa within the Patescibacteria (TM7) phylum often exhibit significant naming variation between pipelines. Full lineage updates are performed to align these entries with the most current taxonomic framework.

```{r}
#| label: tax-harmonization-lineage

# Define full lineage replacement rules for specific TM7 taxa
lineage_correction_map <- tibble::tribble(
  ~species_original,         ~phylum_new,     ~class_new,        ~order_new,          ~family_new,        ~genus_new,      ~species_new,
  "TM7 phylum sp. HMT-352", "Patescibacteria", "Saccharimonadia", "Saccharimonadales", "Nanosynbacteraceae", "Nanosynbacter", "sp. HMT-352",
  "TM7 phylum sp. HMT-346", "Patescibacteria", "Saccharimonadia", "Saccharimonadales", "Saccharimonadaceae", "Saccharimonas", "sp. HMT-346",
  "TM7 phylum sp. HMT-347", "Patescibacteria", "Saccharimonadia", "Saccharimonadales", "Saccharimonadaceae", "Saccharimonas", "sp. HMT-347",
  "TM7 phylum sp. HMT-349", "Patescibacteria", "Saccharimonadia", "Saccharimonadales", "Saccharimonadaceae", "Saccharimonas", "sp. HMT-349",
  "TM7 phylum sp. HMT-351", "Patescibacteria", "Saccharimonadia", "Saccharimonadales", "Nanosyncoccaceae", "Nanosyncoccus", "sp. HMT-351",
  "TM7 phylum sp. HMT-348", "Patescibacteria", "Saccharimonadia", "Saccharimonadales", "Parvisynbacteraceae", "Parvisynbacter", "sp. HMT-348",
  "TM7 bacterium human oral taxon HOT-869", "Patescibacteria", "Saccharimonadia", "Saccharimonadales", "Saccharimonadaceae", "Saccharimonas", "sp. HMT-869",
  "TM7 phylum sp. HMT-356", "Patescibacteria", "Saccharimonadia", "Saccharimonadales", "Nanoperiomorbaceae", "Nanoperiomorbus", "periodonticus",
)

# Apply lineage corrections to the Nanopore dataset via left join
nanopore_processed <- nanopore_processed %>%
  dplyr::left_join(lineage_correction_map, by = c("species" = "species_original")) %>%
  dplyr::mutate(
    phylum  = coalesce(phylum_new, phylum),
    class   = coalesce(class_new, class),
    order   = coalesce(order_new, order),
    family  = coalesce(family_new, family),
    genus   = coalesce(genus_new, genus),
    species = coalesce(species_new, species)
  ) %>%
  dplyr::select(-ends_with("_new"))
```

## Data Export

Finally, the cleaned and harmonized datasets are saved as RDS files to facilitate seamless integration into subsequent analysis modules.

```{r}
#| label: data-export

# Create the output directory if it does not already exist
if (!dir.exists("processed_data")) {
  dir.create("processed_data")
}

# Save serialized data objects
saveRDS(illumina_processed, file = "processed_data/illumina_processed.rds")
saveRDS(nanopore_processed, file = "processed_data/nanopore_processed.rds")
```

The data preprocessing phase is now complete. The harmonized datasets are prepared for comparative diversity analysis and platform evaluation.
